---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: yes
---


## Loading and preprocessing the data

First, loading the zip file and unziping it, then reading the csv data file: 

```{r loading, warning=FALSE , message=F}

if (!file.exists("raw_data.zip")){
  print("loading the zip file............... ")
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileURL, "raw_data.zip", method="curl")
}  
if (!file.exists("activity.csv")) { 
  print("unziping the data...................")
  unzip("raw_data.zip") 
}

library(readr)
activity_full <- read_csv("activity.csv")
library(knitr)
kable(head(activity_full), align='c')
```

Second, processing the data; I delete the lines with NA values:

```{r pressure}
activity <- activity_full[ !is.na(activity_full$steps),]
kable( head(activity), align='c' )
```






## What is mean total number of steps taken per day?

I calcute down below the total number of steps taken per day:

```{r}
activity_sum <- aggregate(steps~date, data=activity, sum)
head(activity_sum)
```

Then I plot the histogram of the total number of steps taken each day


```{r histogram, warning=F}
library(ggplot2)
qplot(steps, data = activity_sum, bins=50)
```


The next step is to calculate the mean and median of the total number of steps taken per day

```{r}
mean_steps <- mean(activity_sum$steps)
median_steps <- round(median(activity_sum$steps),2)

```
The mean of total steps is `r mean_steps`, and their median is `r median_steps`.


```{r}
averages1 <- c( round( mean(activity_sum$steps), 2), round( median(activity_sum$steps), 2))
averages1

```
The mean of total steps is **10766.19** and their median is **10765**.


```{r mean_and_median, eval=FALSE, echo=F, results='hide'}
activity_mean <- aggregate(steps~date, data=activity, mean)
activity_median <- aggregate(steps~date, data=activity, median)
activity_mean
library(knitr)
kable(activity_mean, digits=2, align='c', padding=2)
qplot(date, steps, data = activity_mean, )
p1 <- ggplot(data=activity_mean, aes(x=date, y=steps)) +
      geom_bar(stat="identity", fill="steelblue")+
      ggtitle("Mean of steps taken each day") +
      theme_minimal()
p1
```







  
    
      
       


## What is the average daily activity pattern?

We need to plot the average daily activity pattern corresponding to each 5-minute interval: (the average calculated using the mean function)

```{r}
interval_avrg <- aggregate(steps~interval, data=activity, mean)
p1 <- ggplot(data=interval_avrg, aes(x=interval, y=steps)) +
      geom_line()+
      ggtitle("The average daily activity pattern") +
      xlab(" 5-minute intervals") + ylab("average steps")+
      theme_minimal()
p1
```

Let's get the 5-minute interval which, on average across all the days in the dataset, contains the maximum number of steps:

```{r}
int_max <- interval_avrg[which.max(interval_avrg$steps),]
kable(int_max, digits=2, align='c')
```
  
It's the `r int_max[1,1]` 5-minute interval!






## Imputing missing values

We'll calculate the number of rows that contains NA's values. To do so, we can use the summary function:


```{r}
kable(summary(activity_full), align='c')
```

It's only the `steps` column that has missing value and it has **2304** missing rows.   
  
  
  Since many rows from the first day are missing, my strategy to replace the NAs will be to use the mean of the corresponding 5-minutes-interval across all the days.
  The dataframe `activity_fill_na` is a copy of the original full raw data, but is going to be modified in order to replace the NA values from the `steps` column:




```{r}
nrows <- dim(activity_full)[1]
activity_fill_na <- data.frame(activity_full)
for( i in 1:nrows){
  if ( is.na(activity_fill_na$steps[i])){
    index_mean_intrvl <- which(interval_avrg$interval==activity_fill_na$interval[i])
    activity_fill_na$steps[i] <- interval_avrg$steps[index_mean_intrvl]
  }
}
kable( head(activity_fill_na, 10), align='c')

```

Now let's plot a histogram of the total number of steps taken each day:




```{r}
activity_sum2 <- aggregate(steps~date, data=activity_fill_na, sum)
qplot(steps, data = activity_sum2, bins=50)

```


```{r}
averages2 <- c( round( mean(activity_sum2$steps), 2), round( median(activity_sum2$steps), 2))
averages2

```
The mean of total steps is **10766.19** and their median is **10766.19**.  

**Comparison:** The value of the mean is the same before and after imputing missing values, while the median changed and became equal to the mean. 


## Are there differences in activity patterns between weekdays and weekends?  
For this part I use the dataset with the filled-in missing values and create a new factor variable `days` in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.  
Note that in my r version, the language is french, and so I had to use the french translation for the weekdays.



```{r}
weekdays_FR <- c("lundi", "mardi", "mercredi", "jeudi", "vendredi")
activity_fill_na$days <- factor(ifelse(weekdays(activity_fill_na$date) %in% weekdays_FR, "weekday", "weekend")) 

```


Now I create two ggplots each one corresponding to weekdays or weekends, and then plotting them together using the `ggarrange` function from the `ggpubr` package:



```{r, warning=F, message=F}
activity_wkdays <- subset(activity_fill_na, days=='weekday')
interval_avrg_wkdays <- aggregate(steps~interval, data=activity_wkdays, mean)
pwd <- ggplot(data=interval_avrg_wkdays, aes(x=interval, y=steps)) +
      geom_line()+
      ggtitle("The average daily activity pattern\n through weekdays") +
      xlab(" 5-minute intervals") + ylab("average steps")+
      theme_minimal()

activity_wkends <- subset(activity_fill_na, days=='weekend')
interval_avrg_wkends <- aggregate(steps~interval, data=activity_wkends, mean)
pwe <- ggplot(data=interval_avrg_wkends, aes(x=interval, y=steps)) +
      geom_line()+
      ggtitle("The average daily activity pattern\n through weekends") +
      xlab(" 5-minute intervals") + ylab("average steps")+
      theme_minimal()

library(ggpubr)
ggarrange(pwe, pwd,nrow = 2, ncol = 1)
```

